Manuscript Figures

### 2021-08-26 src/manuscript_figures.Rmd
merge figure 1 and 2 (cancer type prediction + unsupervised clustering benchmarks)
- move cancer type associated LFs plot to supplements
- move prediction accuracy bar plot to supplements
- make a new confusion matrix heatmap in supplements

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors 
res <- sapply(simplify = F, names(LFs), function(tool) {
  message(date(), " => ", tool)
  M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut
  ct <- gsub("-", "_", surv[match(rownames(M), bcr_patient_barcode)]$project)
  res <- predict_labels_glmnet_multiclass(M, ct, partition = 0.6, nodes = 20)
})

@@

dt <- data.table(melt(pancancer_stats))[Var1 %in% c('Mean_Balanced_Accuracy')]
dt$Var2 <- toupper(gsub("_factors", "", dt$Var2))
p1 <- ggplot(dt, aes(x = Var2, y = value)) + 
  geom_bar(stat = 'identity', aes(fill = Var2)) + 
  labs(y = 'Mean Balanced Accuracy') + 
  theme(legend.title = element_blank(), axis.title.x = element_blank(), 
        axis.text = element_text(size = 14), axis.title = element_text(size = 16))

# performance visualisation
#m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
#p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
#         display_numbers = T, number_format = "%.1f", silent = T)

@@

p2 <- plot_tsne(M, ct, show.labels = T) + 
  theme(legend.position = 'none')

p3 <- plot_label_specific_vars(M, ct)

# data type contribution to top LF per factor
p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                           top_lfs = 1, top_features = 100) 
p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'top', legend.background = element_blank(), 
        legend.title = element_blank())

Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B'),
                           rel_widths = c(1,2)), 
                  plot_grid(p4, p3, nrow = 1, labels = c('C', 'D'), 
                            rel_widths = c(1,2)), 
                  nrow = 2,
                  rel_heights = c(2,3))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),
              plot = Fig, units = 'in', width = 12, height = 12)
Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types. A) comparing MAUI with PCA/MOFA/MCIA. Mean Balanced Accuracy computed on the test dataset (30% of held-out data) trained on 70% of the dataset using an Elastic Net multi-class model. B) tSNE plot of MAUI latent factors learned from pan-cancer dataset  C) Percentage of top feature types contributing to top predictive maui latent factor per cancer type. D) Top maui LFs detected per cancer type. Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."

```

```{r clustering_metrics, fig.width=10, fig.height=8}
# clustering metrics of various cancers 
# Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      # to avoid highly correlated maui LFs to dominate clusters
      M <- scale(t(remove_redundant_variables(t(M))))
    } 
    # try topN from 5 to 25 (each tool may capture different amount of information 
    # by varying number of factors)
    stats <- do.call(rbind, lapply(seq(5, min(20, ncol(M)), 5), function(n) {
      message(n)
      M.sub <- t(subset_features_by_variance(t(M), topN = n))
      val <- c('internal') 
      if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
      res <- clValid::clValid(obj = M.sub,
                            nClust = 3:8,
                            clMethods = 'kmeans',
                            validation = val,
                            maxitems = 10000,
                            nstart = 1000, iter.max = 30)
      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
      colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
      stats$top_factors <- n
      return(stats)
    }))
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

@@

p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb,
                                             .SD[which.min(score)],
                                             by = c('cancer_type', 'method', 'k')]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE,  
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), nrow = 1),
                 p4, nrow = 2, labels = c('', 'C'))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.cohort_based_metrics.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),        
       plot = Fig, units = 'in', 
       width = 10, height = 10)
Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B). Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) for each tool. C-D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival (in C) or Progression Free Survival endpoints. (in D)."
```

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 8.27, height = 11.69, units = 'in')
Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction. A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs. B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender). C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."

@@

# print p3 to supplements 
supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- Fig
fig.name <- paste0('SuppFigure', supFig_count, '.adjusted_survival_curves.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 8.27, height = 11.69)
SuppLegends[[supFig_count]] <- "Survival curves adjusted for tumor stage + age + gender based on a cutpoint of 
top survival-associated LF per cancer cohort."
```

>>>>

```{r pancancer_prediction, fig.height=9, fig.width=12}

# In this section, we make figures about pan-cancer classification 
# and some benchmarks of cohort-based unsupervised clustering across cancers

# 1. pancancer classification
# Find out if cancer types are predictable using latent factors 
res <- sapply(simplify = F, names(LFs), function(tool) {
  message(date(), " => ", tool)
  M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut
  ct <- gsub("-", "_", surv[match(rownames(M), bcr_patient_barcode)]$project)
  res <- predict_labels_glmnet_multiclass(M, ct, partition = 0.6, nodes = 20)
})

@@

dt <- data.table(melt(pancancer_stats))[Var1 %in% c('Mean_Balanced_Accuracy')]
dt$Var2 <- toupper(gsub("_factors", "", dt$Var2))
p1 <- ggplot(dt, aes(x = Var2, y = value)) + 
  geom_bar(stat = 'identity', aes(fill = Var2)) + 
  labs(y = 'Mean Balanced Accuracy') + 
  theme(legend.position = 'none', axis.title.x = element_blank(), 
        axis.text = element_text(size = 14), axis.title = element_text(size = 16)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1))

# performance visualisation - misclassified examples 
m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
p2 <- pheatmap::pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
          silent = T, cellwidth = 10, cellheight = 10)

@@

p3 <- plot_tsne(M, ct, show.labels = T) + 
  theme(legend.position = 'none')

p4 <- plot_label_specific_vars(M, ct)

# data type contribution to top LF per factor
p5 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                           top_lfs = 1, top_features = 100) 
p5 <- p5 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'top', legend.background = element_blank(), 
        legend.title = element_blank())

# 2. Cohort-based unsupservised clustering benchmark
# clustering metrics of various cancers 
# Use top most-variable factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      # to avoid highly correlated maui LFs to dominate clusters
      M <- scale(t(remove_redundant_variables(t(M))))
    } 
    # use top 5 most-variable factors for clustering
    M.sub <- t(subset_features_by_variance(t(M), topN = 5))
    val <- c('internal') 
    if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
    cl_res <- clValid::clValid(obj = M.sub,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000, iter.max = 30)
    stats <- data.table::as.data.table(cl_res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

@@

p6 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = comps) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb,
                                             .SD[which.min(score)],
                                             by = c('cancer_type', 'method', 'k')]
dt2$method <- toupper(gsub("_factors", "", dt2$method))
comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
p7 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE,  
                     comparisons = comps) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine plots for main figure 
Fig <- plot_grid(plot_grid(p3, p5, nrow = 1, labels = c('A', 'B'),
                           rel_widths = c(2,1)), 
                  plot_grid(p6, p7, nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,1))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.benchmarks.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),
              plot = Fig, units = 'in', width = 12, height = 12)
Legends[[fig_count]] <- " A) tSNE plot of MAUI latent factors learned from pan-cancer dataset  B) Percentage of top feature types contributing to top predictive maui latent factor per cancer type. C, D) Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in C and cluster stability (Average Proportion of Non-Overlap (APN) in D. Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) for each tool."

# supplementary figure

supFig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, nrow = 1, 
                                                labels = c('A', 'B'), rel_widths = c(1.2, 2)), 
                          p4, nrow = 2, labels = c('', 'C'), rel_heights = c(1.2, 2))
# print p3 to supplements 
supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- supFig
fig.name <- paste0('SuppFigure', supFig_count, '.benchmarks.supp.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = supFig, units = 'in', width = 8.27, height = 11.69)
SuppLegends[[supFig_count]] <- "A) Predictive power of latent factors in distinguishing cancer types comparing MAUI with PCA/MOFA/MCIA. Mean Balanced Accuracy computed on the test dataset (30% of held-out data) trained on 70% of the dataset using an Elastic Net multi-class model. B) Confusion matrix of predicted versus true labels of samples' cancer types. The color of each cell represents the percentage of samples from the cancer type in the rows that are classified into the cancer types on the column. Cells that deviate from the red color represent relatively lower
level of classification accuracy. C) Top maui LFs detected per cancer type. Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
```

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 8.27, height = 11.69, units = 'in')
Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type. B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs."

@@

# print p3 to supplements 
supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- Fig
fig.name <- paste0('SuppFigure', supFig_count, '.adjusted_survival_curves.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 8.27, height = 11.69)
SuppLegends[[supFig_count]] <- "Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
```
